Evaluating the Capabilities of the Lattice Boltzmann Method for non-Newtonian and Free-surface Flow towards Applications in Wellbore Cementing

### 2016-09-02 src/stability-paper.tex
revised figures and addressed comments for publication

The Lattice Boltzmann method (LBM) is a method based on computational statistical mechanics that is well-suited for complex flow such as non-Newtonian, free surface, and multiphase multicomponent flow.
Non-Newtonian flow is the primary focus of this Chapter, as many practical engineering problems such as the flow of cement slurry and concrete, the filling of molds by molten metals and plastics, blood flow, etc., are best modeled as non-Newtonian fluids.
LBM is typically applied to simulate flow through a series of time seps, each consisting of streaming particle distributions to neighboring nodes, and collisions of particle distributions at each node through a collision operator.
The collision operator is of interest because it, along with the equilibrium distribution function, determine the physics that are simulated, e.g. constitutive laws, interfacial dynamics, etc., and it has implications on numerical stability and computational efficiency.
In this Chapter, various collision operators and methods for stability enhancement were examined for their suitability for simulating non-Newtonian fluid flows in terms of accuracy, numerical stability and computational efficiency.
The investigation was carried out as a numerical study looking for qualitative, yet practical, results; it included testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluids.
Two different benchmark problems were chosen for the flows: Poiseuille flow, and lid-driven square cavity flow.
The test results are followed by recommendations for choice collision scheme given priority and problem type.

@@

The dynamic viscosity of a fluid is a measure of its resistance to shear deformation.
For many fluids, at a constant temperature, the dynamic viscosity can be considered as constant.
These fluids are known as Newtonian fluids.
Non-Newtonian fluids have an apparent dynamic viscosity, or apparent resistance to shear stress, that is variable even at a constant temperature and is often a function of strain-rate.
There are a number of fluids in science and engineering applications that can be classified as non-Newtonian: pastes, slurries, molten plastics, polymer solutions, dyes, varnishes, suspensions, and some biomedical liquids such as blood all behave in a non-Newtonian manner~\cite{bohme1987non}.

Of all of the different non-Newtonian behaviors that exist, there are two models under which much of the behaviors may be idealized: yield stress fluids and power-law fluids.
Yield stress fluids, also known as Bingham plastics, do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
Yield stress flow is relevant in many other disciplines and applications because of the many substances that exhibit yield stress behavior, e.g. pastes, paints, muds, molten plastics and metals, and in some cases, blood~\cite{wang2011lattice}. %MG: citation is for blood as non-Newtonian
As a consequence, simulating yield stress fluids can help engineers develop better pastes and paints, understand the flow and deformation of mud and clay in geotechnical engineering, understand blood circulation, and better manufacture metal and plastic parts.

There other idealized non-Newtonian behavir, power-law behavior, is more commonly known as \emph{shear-thinning}--when the apparent viscosity decreases with increasing strain-rate--or \emph{shear-thickening}--when the apparent viscosity increases with increasing strain-rate. Shear-thinning fluids are also known as \emph{pseudoplastics}, and some examples include polymer mixes and molten plastics.
Shear-thickening fluids are also known as \emph{dilatants}, and some examples include quicksand, a cornstarch and water mixture, and a silica and polyethylene glycol mixture.

@@

Analytical solutions rarely exist for even the simplest non-Newtonian fluid flows because of the complexity that a nonlinear constitutive relationship entails.
It is generally more practical to approximate the flow of non-Newtonian fluids using numerical methods.
However, the nonlinear constitutive equation--typically of the form \begin{equation} \label{eq:non-newtonian}
\tau = \mu_{app}(\dot{\gamma}) \dot{\gamma},
\end{equation}\noindent where the apparent viscosity, $\mu_{app}$ is a function of the strain rate--results in certain challenges for numerical methods as well.
Determining the apparent viscosity and strain-rate distribution of a flow over time will often require an iterative solution at every time step, a general Picard-type algorithm for such a process can be given as:

\begin{enumerate}
	\item Start with initial guess for the apparent viscosity, $\mu_k = \mu_0$.
	\item \label{step:solve-for-flow} Solve for the flow using the current value of apparent viscosity, $\dot{\gamma}_k = \tau_k / \mu_k$.
	\item Update the apparent viscosity, $\mu_{k+1} = \mu(\dot{\gamma}_k)$.
	\item Return to Step \ref{step:solve-for-flow} until convergence is met.
\end{enumerate}

Numerical solutions work by discretizing the equations that govern the physics of interest.
When the solution for fluid flow problems vary in space and time, a numerical approximation requires breaking the problem up into discrete locations and discrete time steps.
The significance of approximating a solution by discretizing the governing equations is that the iterative solution for the constitutive equation must be solved at each discrete location for each discrete time step, which can become computationally expensive.
The lattice Boltzmann method (LBM) is a numerical method for fluid flow that has the advantage that computing the strain rate is second-order accurate in space and local to each node~\cite{kruger2010second}.
This means that although an iterative solution is still required to determine the local strain rate and apparent viscosity, each iterative solution can be done in parallel, by a separate process, as they are independent of each other.
Because hardware architectures have shifted from single, sequential processing systems to parallel processing systems, the local nature of the stress--strain-rate relationship in LBM gives it a distinct advantage for simulating non-Newtonian flow over some other numerical methods.

The lattice Boltzmann method has been studied and successively applied to modeling various flows of non-Newtonian fluids.
For example, ~\citet{tang2011bingham,chai2011multiple,fallah2012multiple,chen2014simulations,vikhansky2008lattice,wang2008lattice} developed LBM models for simulating yield stress flow.
The LBM model results agreed well when compared to analytical solutions for Bingham plastic Poiseuille flow and values from literature for lid-driven cavity flow, which shows the feasibility of using LBM models for yield-stress fluids.
LBM models for power-law fluid flow~\citet{wang2011lattice,wang2015localized,boyd2006second,chai2011multiple}, and blood flows using the K-L, Casson, and Carreau-Yasuda constitutive relationships~\citet{ashrafizaadeh2009comparison}, have also been successfully developed and verified. %MG: does citing this way appear too dense?

@@

One such consideration associated with this view is the potential for numerical inaccuracies and instabilities ~\cite{sterling1993stability,sterling1996stability,bawazeer2013stability,lallemand2000theory}. %MG: are all relevant sources cited?

@@

Various schemes and strategies for incorporating the physics of non-Newtonian flow with LBM and yet maintaining a stable numerical method have been developed and studied.
The simplest approach for simulating a shear-rate dependent viscosity, used in~\cite{boyd2006second,chen2014simulations,fallah2012multiple,tang2011bingham,svec2011flow,svec2012free,zhao2016lattice}, is to make the collision frequency, which is proportional to apparent viscosity, variable and dependent on the local strain rate.
A potential issue with the stability of the variable relaxation time approach is that as the collision frequency approaches $2$ the viscosity approaches zero and overrelaxation occurs; and alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method also degrades~\cite{latt2007hydrodynamic}. %MG: verify this statement

@@

Although bounding the collision frequency was shown to be effective in terms of stability, it is also nonphysical and can lead to approximations that are inaccurate, not because of round-off error or numerical instability, but because the collision does not reflect the proper constitutive relationship of interest.

@@

The equilibrium distribution function is derived for the specific constitutive relationship of interest using the Chapman-Enskog multiscale expansion.

@@

The strategy of using an equilibrium distribution function that incorporates the local shear-rate effect has the advantage that, because the collision frequency is constant (at unity), the collision frequency will not approach zero, and will not result in overrelaxation, and the collision frequency will not reach values much greater than one, and so there is no reason to bound the collision frequency in a way that is nonphysical.
\citet{wang2011lattice} developed another constant-collision frequency LBM scheme for non-Newtonian flow by splitting the effects of constitutive relationship into Newtonian and non-Newtonian parts.
The Newtonian part was modeled in the usual way, namely scaling the collision frequency to achieve the macroscopic (Newtonian) viscosity, and the non-Newtonian part was modeled as a source of momentum, i.e. as an external forcing term, that is dependent on local shear-rate.
Although the constant-collision frequency strategies present interesting alternatives, the variable collision scheme is used in the present study because of its simplicity and generality, i.e. it can be easily fitted to any constitutive equation without for example, performing Chapman-Enskog multiscale expansion.

To improve upon the stability of LBM models,~\citet{d1994generalized} developed a multiple-relaxation-time (MRT) collision operator, that takes place in moment space and allows each moment to relax at a different rate.
\citet{lallemand2000theory} used von Neumann stability analysis to investigate the stability of the newly constructed LBM-MRT model and concluded LBM with the MRT collision operator was more stable, but with increased computational expense than the commonly used BGK collision operator.
Note that although this increased computational expense was decided to not be significant for Newtonian flows ($\approx$ 10-20\%~\cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flow because an iterative solution for the constitutive equation can require that certain expensive computations be performed at each iteration (the strain rate tensor is determined from the nonequilibrium particle distribution which must be mapped into moment space when using the MRT collision operator).
\citeauthor{chen2014simulations} concluded that the MRT collision operator was more stable for Bingham plastic flow and allowed the use of a more accurate approximation to the Bingham plastic constitutive relationship.
However, what remains unclear is:
\begin{itemize}
    \item What is the increased cost associated with the MRT collision operator when applied to non-Newtonian flow?
    \item Under what conditions, e.g. material parameters, physical problem, etc., for Bingham plastic fluid flow is the MRT collision operator necessary to maintain stability and/or accuracy?
    \item Under what conditions, e.g. material parameters, physical problem, etc, for power-law fluid flow is the MRT collision operator necessary to maintain stability and/or accuracy?
    \item What are additional strategies for increasing stability and accuracy, and what are their associated computational costs?
\end{itemize}
In regards to the last question, much work has been done recently to enhance stability of LBM models beyond the MRT collision operator.
\citet{brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy,gorban2014enhancement} have all developed and tested means for introducing artificial dissipation in order to dampen out high frequency, nonphysical oscillations.
Stability enhancement through artificial dissipation and entropic filtering has shown a lot of promise, but to the authors' knowledge has not been tested for use in simulating the flow of non-Newtonian fluids.

The goal of this Chapter is to numerically study the implications of accuracy, stability, and efficiency for some of the different strategies for simulating non-Newtonian flow using LBM. The intention of the study is to aid scientists and engineers in understanding which strategy is best suited to their priorities and applications of interest so as to maximize the advantages LBM has in simulating non-Newtonian flow.
Advantages, such as LBM's potential to scale well in parallel, can be much less realized if the collision operator is too computationally expensive, or if numerical instabilities ensue.
A numerical study can help to determine approximate numerical values, domains, and boundary conditions in which one LBM scheme may be more advantageous than another so that LBM may be used in a computationally efficient and stable manner.

@@

LBM has some advantages over other methods of CFD.
For example, LBM is a computationally efficient approach for some CFD.
This efficiency is a consequence of two distinct features of LBM: (1) the convective operator is linear, as opposed to the nonlinear convection terms that appear in continuum mechanics approaches; and (2) the fluid pressure is given by an equation of state.
Solving for the fluid pressure in traditional method is more computationally expensive and requires special treatment such as iteration and/or relaxation~\cite{chen1998lattice}.

@@

\begin{itemize}
\item Streaming: particle distributions propagate to their neighbors along the lattice velocity vectors.
The particles can only move along the vectors in their specified direction and can only move at a specific speed.
\item Collision: particle distributions meet at a node and ``collide''.
In LBM, collisions are not simulated in a realistic sense, meaning that each individual particle does not exist and glance off of, or interact with, one another.
Instead the collision operator is formulated in such a way that particle distributions are relaxed toward equilibrium.
What defines equilibrium depends on the mechanics of interest to be modeled.
\end{itemize}

@@

\pagebreak
The macroscopic variables of interest can be calculated from the particle distribution functions, $f(\pos, \pvel, t)$, by integrating moments of $f$ over velocity space.
Due to the discrete nature of velocity in LBM, the integrals simply become summations.
The mass density is given by the sum of the particle distributions and the momentum density is given by the first moment of the particle distributions over the velocity space:

@@

\subsubsection{Bhatnagar-Gross-Krook (BGK)} \label{sec:bgk}

The collision operator, in the case of the continuous BE, attempts to describe the change in particle momentums and trajectories due to pairwise particle collisions (based on their respective momentums and trajectories just prior to collision)~\cite{Cer90}.
In LBM, the collision operator causes particle distributions to relax toward a quasiequilibrium. % MG: is this true of all collision operators? or just BGK?
This equilibrium is determined by the macroscopic physical behavior of interest.
In the case of incompressible flow, the quasiequilibrium particle distribution, $f_i^{eq} = f_i^{eq}(\pos, t)$, is often given by:

@@

\noindent where $\rho$ and $\mvel$ are dependent on $\pos$ and $t$, and $w_i$ is the weight in the $i^{th}$:%MG: maybe say something about where this comes from, e.g. feq is a Hermite polynomial expansion of the Maxwell-Boltzmann distribution

@@

\noindent where $\omega$ is the collision frequency~\cite{Bha54}.
The collision frequency can be related to macroscopic constitutive properties through Chapman-Enskog multiscale analysis~\cite{wolf2000lattice}.
For incompressible Newtonian flow, the collision frequency is related to the kinematic viscosity by $\nu = c_s^2(\frac{1}{\omega} - \frac{1}{2})$; and from this relationship it is clear that $\omega \in [0.0, 2.0]$, otherwise the viscosity would be negative.
The method for simulating non-Newtonian flow in the current work involves approximating a solution to the local value of apparent viscosity, $\mu_{app}(\pos, t)$, using \eqref{eq:non-newtonian} where $\tau$ and $\dot{\gamma}$ are dependent on $\pos$ and $t$, and then setting the value of the local collision frequency as follows:
\begin{equation} \label{eq:colfeq-non-newtonian}
\omega(\pos, t) = \frac{1}{\frac{\mu_{app}(\pos, t)}{c_s^2 \rho(\pos, t)} + \frac{1}{2}}
\end{equation}

Despite the utility of the BGK collision operator, it does have a few drawbacks.
For example, in low viscosity fluids the BGK operator results in an overrelaxation of particle distributions toward quasiequilibrium.
It is well known that when large nonequilibrium distributions exist in the LBM approximation that overrelaxation can result in nonphysical oscillations that are slow to decay~\cite{brownlee2007stability,dellar2003incompressible}.
To illustrate this, consider a flow in which there is a sharp spatial gradient in either $\rho$ or $\mvel$ at $\pos$.
As $f_i^{eq}$ depends on both $\rho$ and $\mvel$ \eqref{eq:feq}, it may be the case that $|f_i^{eq}(\pos, t) - f_i^{eq}(\pos + \pvel_i \Delta t, t + \Delta t)| >> 0$, i.e. there may be a large difference in the quasiequilibrium for the $i^{th}$ discrete velocity at $(\pos, t)$ and $(\pos + \pvel_i \Delta t, t + \Delta t)$.
In this case, if $f_i$ is ``near'' to quasiequilibrium at $\pos$ it will be ``far'' from quasiequilibrium after the streaming step when it moves to the node at $\pos + \pvel_i \Delta t$.
Overrelaxation will occur if $\nu \rightarrow 0$ because consquently $\omega \rightarrow 2$ and \eqref{eq:bgk} results in $f_i(\pos + \pvel_i \Delta t,t + \Delta t)$ still being ``far'' from $f^{eq}_i(\pos + \pvel_i \Delta t,t + \Delta t)$ but on the ``other side''.
Overrelaxation in subsequent time steps (along the streaming trajectory of $f_i$) could result in nonphysical oscillations such as those depicted in \Fref{fig:overrelaxation}.
Considering the effect oscillations of particle distributions will have on macroscopic variables and, consequently, local quasiequilibriums, positive feedback loops can occur causing the system to diverge or ``pollute'' the system enough to make the results highly nonphysical and altogether useless~\cite{gorban2014enhancement}.

\begin{figure}
	\centering
    \includegraphics[width=\linewidth]{figs/overrelaxation_tau-0513}
    \caption{Example of oscillations that can occur as a result of overrelaxation; the collision frequency, $\omega = 1.95$ and $\mu_{app} \approx 0.0043$}
    \label{fig:overrelaxation}
\end{figure}

The challenge associated with high viscosity fluids is that particular distributions may never relax as ``far'' toward quasiequilibrium as is physical because $\omega \rightarrow 0$ as $\nu \rightarrow \infty$, resulting in extreme underrelaxation to the point of being negligible. %MG: this needs cleaned up, I think

Concerns with sharp gradients, overrelaxation, and underrelaxation are particularly relevant in non-Newtonian flow because of the nonlinear constitutive relationship between shear stress and strain-rate.
The nonlinear constitutive relationship can lead to sharp gradients in $\rho$ or $\mvel$, and depending on the form of the function $\mu_{app}(\dot{\gamma})$, the apparent viscosity, $\mu(\dot{\gamma}(\pos))$, may result in overrelaxation in certain parts of the domain and extreme underrelaxation in others.

@@

\subsubsection{Multiple-relaxation Time (MRT)}

An alternative to the BGK collision operator is the multiple-relaxation-time (MRT) collision operator.
In the LB-MRT scheme, one constructs a space based on the particle velocity, $\pvel$, moments of $\mathbf{f} = \begin{Bmatrix}f_0 & f_1 & ... & f_8\end{Bmatrix}$, herein referred to as the ``moment space''.
The collision is then performed in the moment space.
There are a few reasons why it is advantageous to perform the collision in the moment space as opposed to the particle distribution space:
\begin{enumerate}
    \item Physical processes in fluids can be approximately described by coupling or interacting among modes, and the modes are directly related to the moments~\cite{lallemand2000theory}.
    \item For the D2Q9 lattice, there are nine distribution functions, $f_0, f_1, ..., f_8$, but only six variables that affect the intended hydrodynamics on a macroscopic scale, namely: $\rho$, $\mathbf{u}$, and $\mathbf{\Pi}$, where $\mathbf{\Pi}$ is the momentum flux tensor~\cite{dellar2003incompressible}.
    Of the nine relaxation rates available, the three that correspond to the extra variables--often referred to as ``ghost variables'', and their associated modes as ``ghost modes''--can be tuned in order to dampen out their associated ghost modes, ensuring these modes do not dominate or cause numerical instabilities at the lattice scale.
\end{enumerate}

@@

To answer the first question, consider the discussion in \Fref{sec:bgk} on problems with overrelaxation and nonphysical oscillations.
Nonphysical oscillations due to overrelaxation would be damped out more quickly if particle distributions ``far'' from quasiequilibrium were brought closer to quasiequilibrium.
A particle distribution vector, $\mathbf{f}$, ``far'' from equilibrium would be a good candidate for artificial dissipation.
There are many ways one can measure the distance between $\mathbf{f}$ and $\mathbf{f}^{eq}$; for example, a reasonable choice would be $||\mathbf{f} - \mathbf{f}^{eq}||_p$ for some $p$ norm.
A metric that has been developed and used successfully for determining when artificial dissipation should be introduced at a lattice site is the so called relative nonequilibrium entropy~\cite{gorban2014enhancement,brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy}.
The relative nonequilibrium entropy, $\Delta S$, is given by:

>>>>

The Lattice Boltzmann method (LBM) is a numerical method based on computational statistical mechanics that is well-suited for approximating complex flow behaviors such as non-Newtonian, free surface, and multiphase multicomponent flow.
LBM is typically applied to simulate flow through a series of time steps, each consisting of streaming particle distributions to neighboring nodes and collisions of particle distributions at each node through a collision operator.
The collision operator is of interest because it, along with the equilibrium distribution function, determine the physics that are simulated (e.g. constitutive laws, interfacial dynamics, etc.) and it has implications on numerical stability and computational efficiency.
This work examines various collision operators and methods for stability enhancement for their suitability for simulating non-Newtonian fluid flows in terms of accuracy, numerical stability and computational efficiency.
The investigation was carried out as a numerical study looking for qualitative, yet practical, results, including testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluids.
Two different benchmark problems were chosen for the flows: Poiseuille flow, and lid-driven square cavity flow.
The test results are followed by recommendations for choice collision scheme given priority and problem type.

@@

There are a number of fluids in science and engineering applications that can be classified as non-Newtonian, such as pastes, slurries, molten plastics, polymer solutions, dyes, varnishes, suspensions, and some biomedical liquids such as blood all behave in a non-Newtonian manner~\cite{bohme1987non}.
Of all of the different non-Newtonian behaviors that exist, there are two models under which much of the behaviors may be idealized: yield stress fluids and power-law fluids.
Yield stress fluids, also known as Bingham plastics, do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
Yield stress flow has been found to be relevant in many applictions, such as flow of pastes, paints, muds, molten plastics and metals, and in some cases, blood~\cite{wang2011lattice}. %MG: citation is for blood as non-Newtonian

Power-law behavior, is more commonly known as \emph{shear-thinning}, when the apparent viscosity decreases with increasing strain-rate, or \emph{shear-thickening}, when the apparent viscosity increases with increasing strain-rate.
Shear-thinning fluids are also known as \emph{pseudoplastics}, and some examples include tomato concentrate, clay and water mixtures, and molten plastics~\cite{irgens2014rheology}.
Shear-thickening fluids are also known as \emph{dilatants}, and a well-known example is a cornstarch and water mixture.

@@

Analytical solutions rarely exist for even the simplest non-Newtonian fluid flows because of the complexity that a nonlinear constitutive relationship entails.
It is generally more practical to approximate the flow of non-Newtonian fluids using numerical methods~\cite{syrakos2014performance,owens2002computational,chai2011multiple}.

The lattice Boltzmann method (LBM) is a numerical method for simulating fluid flow that has the advantage (among others) that computing the strain-rate is second-order accurate in space and local to each node~\cite{kruger2010second}.
This means that although an iterative solution is still required to determine the local strain-rate and apparent viscosity, each iterative solution can be done in parallel, by a separate process, as they are independent of each other.
Because hardware architectures have shifted from single, sequential processing systems to parallel processing systems, the local nature of the stress--strain-rate relationship in LBM gives it a distinct advantage for simulating non-Newtonian flow over some other numerical methods.
For example,~\cite{tang2011bingham,chai2011multiple,fallah2012multiple,chen2014simulations,vikhansky2008lattice,wang2008lattice} developed LBM models for simulating yield stress flow.
The LBM model results agreed well when compared to analytical solutions for Bingham plastic Poiseuille flow and values from literature for lid-driven cavity flow, which shows the feasibility of using LBM models for yield-stress fluids.
Examples of LBM models for power-law fluid flow include~\cite{wang2011lattice,wang2015localized,boyd2006second,chai2011multiple}, and blood flows using the K-L, Casson, and Carreau-Yasuda constitutive relationships~\cite{ashrafizaadeh2009comparison}, which have also been successfully verified with benchmarks.

@@

One such consideration is the potential for numerical inaccuracies and instabilities ~\cite{sterling1993stability,sterling1996stability,bawazeer2013stability,lallemand2000theory}. %MG: are all relevant sources cited?

@@

Various strategies for incorporating the physics of non-Newtonian flow with LBM while maintaining a stable numerical method have been developed and studied.
The simplest approach for simulating a shear-rate dependent viscosity is to make the collision frequency, which is proportional to apparent viscosity, variable and dependent on the local strain-rate~\cite{boyd2006second,chen2014simulations,fallah2012multiple,tang2011bingham,svec2011flow,svec2012free,zhao2016lattice}.
A potential issue with the stability of the variable relaxation time approach is that as the collision frequency approaches $2$ the viscosity approaches zero and overrelaxation occurs.
Alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method degrades~\cite{latt2007hydrodynamic}. %MG: verify this statement

@@

Although bounding the collision frequency was shown to be effective in terms of stability, it is nonphysical and can lead to approximations that are inaccurate, not because of round-off error or numerical instability, but because the collision does not reflect the proper constitutive relationship of interest.

@@

The equilibrium distribution function can be derived for the specific constitutive relationship of interest using the Chapman-Enskog multiscale expansion.

@@

Using an equilibrium distribution function that incorporates the local shear-rate effect has the advantage that, because the collision frequency is constant (at unity), the collision frequency will not approach values that lead to overrelaxation (e.g. $2$) or underrelaxation (e.g. $\approx 0$).
\cite{wang2011lattice} developed another constant-collision frequency LBM scheme for non-Newtonian flow by splitting the effects of constitutive relationship into Newtonian and non-Newtonian parts.
The Newtonian part was modeled in the usual way, namely scaling the collision frequency to achieve the macroscopic (Newtonian) viscosity, and the non-Newtonian part was modeled as a source of momentum (i.e. as an external forcing term) that is dependent on local shear-rate.
Although the constant-collision frequency strategies present interesting alternatives, the variable collision frequency scheme is used in the present study because of its simplicity and generality, such as the fact that a variable collision frequency scheme can be fit to any constitutive relationship without performing Chapman-Enskog multiscale expansion.

%The Bhatnagar, Gross, Krook (BGK) collision operator~\cite{bhatnagar1954model}.
The Multiple-relaxation-time (MRT) collision operator ~\cite{d1994generalized} is another approach that has been developed to improve upon the stability of LBM.
The MRT collision takes place in moment space and allows each moment to relax at a different rate.
\cite{lallemand2000theory} used von Neumann stability analysis to investigate the stability of the newly constructed LBM-MRT model for fluid flow, and concluded LBM with the MRT collision operator was more stable, but with increased computational expense than the commonly used BGK collision operator.
Note that although this increased computational expense was decided to not be significant for Newtonian flows ($\approx$ 10-20\%~\cite{lallemand2000theory}), the cost may increase significantly for non-Newtonian flow because an iterative solution for the constitutive equation can require that certain expensive computations be performed at each iteration (the strain-rate tensor is determined from the nonequilibrium particle distribution which must be mapped into moment space when using the MRT collision operator).
\cite{chen2014simulations} concluded that the MRT collision operator was more stable for Bingham plastic flow and allowed the use of a more accurate approximation to the Bingham plastic constitutive relationship.
However, what remains unclear is:
\begin{itemize}
    \item What is the increased cost associated with the MRT collision operator when applied to non-Newtonian flow?
    \item Under what conditions (e.g. material parameters, physical problem, etc.) for Bingham plastic fluid flow is the MRT collision operator necessary to maintain stability and/or accuracy?
    \item Under what conditions (e.g. material parameters, physical problem, etc.) for power-law fluid flow is the MRT collision operator necessary to maintain stability and/or accuracy?
    \item What are additional strategies for increasing stability and accuracy, and what are their associated computational costs?
\end{itemize}
In regards to the last question, much work has been done recently to enhance stability of LBM models beyond the MRT collision operator.
\cite{brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy,gorban2014enhancement} have all developed and tested means for introducing artificial dissipation in order to dampen out high frequency, nonphysical oscillations.
Stability enhancement through artificial dissipation and entropic filtering has shown promise, but to the authors' knowledge has not yet been tested for use in simulating the flow of non-Newtonian fluids.

The goal of the present study is to numerically study the implications on accuracy, stability, and efficiency for some of the different strategies for simulating non-Newtonian flow using LBM. The intention of the study is to aid scientists and engineers in understanding which strategy is best suited to their priorities and applications of interest so as to maximize the advantages LBM has in simulating non-Newtonian flow.
Advantages, such as LBM's potential to scale well in parallel, can be much less realized if the collision operator is too computationally expensive, or if numerical instabilities ensue.
A numerical study can help to determine approximate numerical values, domains, and boundary conditions in which one LBM scheme may be more advantageous than another so that LBM may be used in a computationally efficient and stable manner.
In \Fref{sec:LBM}, an overview of the Lattice Boltzmann Method for simulating fluid flow is presented, along with the collision operators, stability enhancements, and constitutive relationships that were investigated.
\Fref{sec:numerical-study} outlines the boundary conditions, material parameters, and LBM collision schemes for each simulation in the numerical study; and is followed by results and discussion.
Finally, concluding remarks are given in \Fref{sec:numerical-study-conclusions}.

@@

As noted previously, LBM has some advantages over other methods of CFD.
For example, LBM is a computationally efficient approach for some CFD.
This efficiency is a consequence of two distinct features of LBM: (1) the convective operator is linear, as opposed to the nonlinear convection terms that appear in continuum mechanics approaches; and (2) the fluid pressure is given by an equation of state.
Solving for the fluid pressure in traditional method is more computationally expensive and requires special treatment such as iteration and/or relaxation~\cite{chen1998lattice}.

@@

\begin{itemize}
\item Streaming: particle distributions propagate to their neighbors along the lattice velocity vectors.
The particles can only move along the vectors in their specified direction and can only move at a specific speed.
\item Collision: particle distributions meet at a node and ``collide''.
In LBM, collisions are not simulated in a realistic sense, meaning that each individual particle does not exist and glance off of or interact with one another.
Instead, the collision operator is formulated in such a way that particle distributions are relaxed toward equilibrium.
What defines equilibrium depends on the mechanics of interest to be modeled.
\end{itemize}

@@

The macroscopic variables of interest can be calculated from the particle distribution functions, $f(\pos, \pvel, t)$, by integrating moments of $f$ over velocity space.
Due to the discrete nature of velocity in LBM, the integrals simply become summations.
The mass density is given by the sum of the particle distributions and the momentum density is given by the first moment of the particle distributions over the velocity space:

@@

\subsubsection{Bhatnagar-Gross-Krook} \label{sec:bgk}

The collision operator, in the case of the continuous BE, attempts to describe the change in particle momentums and trajectories due to pairwise particle collisions (based on their respective momentums and trajectories just prior to collision)~\cite{Cer90}.
In LBM, the collision operator causes particle distributions to relax toward a quasiequilibrium. % MG: is this true of all collision operators? or just BGK?
This equilibrium is determined by the macroscopic physical behavior of interest (e.g. diffusion, incompressible flow, compressible flow, etc.).
In the case of incompressible flow, the quasiequilibrium particle distribution, $f_i^{eq} = f_i^{eq}(\pos, t)$, is often given by:

@@

\noindent where $\rho$ and $\mvel$ are dependent on $\pos$ and $t$, and $w_i$ is the weight in the $i$th:%MG: maybe say something about where this comes from, e.g. feq is a Hermite polynomial expansion of the Maxwell-Boltzmann distribution

@@

\noindent where $\omega$ is the collision frequency~\cite{Bha54}.
The collision frequency can be related to macroscopic constitutive properties through Chapman-Enskog multiscale analysis~\cite{wolf2000lattice}.
For incompressible Newtonian flow, the collision frequency is related to the kinematic viscosity by $\nu = c_s^2(\frac{1}{\omega} - \frac{1}{2})$; and from this relationship it is clear that $\omega \in [0.0, 2.0]$, otherwise the viscosity would be negative.
The method for simulating non-Newtonian flow in the current work involves approximating a solution to the local value of apparent viscosity, $\mu_{app}(\pos, t)$, where the shear stress, $\tau$, and strain-rate, $\dot{\gamma}$, are dependent on $\pos$ and $t$, and then setting the value of the local collision frequency as follows:
\begin{equation} \label{eq:colfeq-non-newtonian}
\omega(\pos, t) = \frac{1}{\frac{\mu_{app}(\pos, t)}{c_s^2 \rho(\pos, t)} + \frac{1}{2}}
\end{equation}

Despite the utility of the BGK collision operator, it does have a few drawbacks.
For example, in low viscosity fluids the BGK operator results in an overrelaxation of particle distributions toward quasiequilibrium.
It is well known that when large nonequilibrium distributions exist in the LBM approximation that overrelaxation can result in nonphysical oscillations that are slow to decay~\cite{brownlee2007stability,dellar2003incompressible}.
To illustrate this, consider a flow in which there is a sharp spatial gradient in either $\rho$ or $\mvel$ at $\pos$.
As $f_i^{eq}$ depends on both $\rho$ and $\mvel$ \eqref{eq:feq}, it may be the case that $|f_i^{eq}(\pos, t) - f_i^{eq}(\pos + \pvel_i \Delta t, t + \Delta t)| >> 0$, i.e. there may be a large difference in the quasiequilibrium for the $i^{th}$ discrete velocity at $(\pos, t)$ and $(\pos + \pvel_i \Delta t, t + \Delta t)$.
In this case, if $f_i$ is ``near'' to quasiequilibrium at $\pos$ it will be ``far'' from quasiequilibrium after the streaming step when it moves to the node at $\pos + \pvel_i \Delta t$.
Overrelaxation will occur if $\nu \rightarrow 0$ because consquently $\omega \rightarrow 2$ and \eqref{eq:bgk} results in $f_i(\pos + \pvel_i \Delta t,t + \Delta t)$ still being ``far'' from $f^{eq}_i(\pos + \pvel_i \Delta t,t + \Delta t)$ but on the ``other side''.
Overrelaxation in subsequent time steps (along the streaming trajectory of $f_i$) could result in nonphysical oscillations.
Considering the effect oscillations of particle distributions will have on macroscopic variables and, consequently, local quasiequilibriums, positive feedback loops can occur causing the system to diverge or ``pollute'' the system enough to make the results highly nonphysical~\cite{gorban2014enhancement}.

The challenge associated with high viscosity fluids is that particular distributions may never relax as ``far'' toward quasiequilibrium as is physical because $\omega \rightarrow 0$ as $\nu \rightarrow \infty$, resulting in extreme underrelaxation to the point of being negligible.
Concerns with sharp gradients, overrelaxation, and underrelaxation are particularly relevant in non-Newtonian flow because of the nonlinear constitutive relationship between shear stress and strain-rate.
The nonlinear constitutive relationship can lead to sharp gradients in $\rho$ or $\mvel$, and depending on the form of the function $\mu_{app}(\dot{\gamma})$, the local collision frequency may result in overrelaxation in certain parts of the domain and extreme underrelaxation in others.

@@

\subsubsection{Multiple-relaxation Time}

An alternative to the BGK collision operator is the multiple-relaxation-time (MRT) collision operator.
In the LB-MRT scheme, one constructs a space based on the particle velocity, $\pvel$, moments of $\mathbf{f} = \begin{Bmatrix}f_0 & f_1 & ... & f_8\end{Bmatrix}^T$, herein referred to as the ``moment space''.
The collision is then performed in the moment space.
There are a few reasons why it is advantageous to perform the collision in the moment space as opposed to the particle distribution space:
\begin{enumerate}
    \item Physical processes in fluids can be approximately described by coupling or interacting among modes, and the modes are directly related to the moments~\cite{lallemand2000theory}.
    \item For the D2Q9 lattice, there are nine distribution functions, $f_0, f_1, ..., f_8$, but only six variables that affect the intended hydrodynamics on a macroscopic scale, namely: $\rho$, $\mathbf{u}$, and $\mathbf{\Pi}$, where $\mathbf{\Pi}$ is the momentum flux tensor~\cite{dellar2003incompressible}.
    Of the nine relaxation rates available, the three that correspond to the extra variables--often referred to as ``ghost variables'', and their associated modes as ``ghost modes''--can be tuned in order to dampen out their associated ghost modes, ensuring these modes do not dominate or cause numerical instabilities at the lattice scale.
\end{enumerate}

@@

%To answer the first question, consider the discussion in \Fref{sec:bgk} on problems with overrelaxation and nonphysical oscillations.
Nonphysical oscillations due to overrelaxation would be damped out more quickly if particle distributions ``far'' from quasiequilibrium were brought closer to quasiequilibrium.
Thus, a particle distribution vector, $\mathbf{f}$, ``far'' from quasiequilibrium would be a good candidate for artificial dissipation.
However, there are many ways one can measure the distance between $\mathbf{f}$ and $\mathbf{f}^{eq}$; for example, a reasonable choice would be $||\mathbf{f} - \mathbf{f}^{eq}||_p$ for some $p$ norm.
A metric that has been developed and used successfully for determining when artificial dissipation should be introduced at a lattice site is the so called relative nonequilibrium entropy~\cite{gorban2014enhancement,brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy}.
The relative nonequilibrium entropy, $\Delta S$, is given by:
